Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-08-02 appendix.tex
Getting closer.

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
    \multicolumn{3}{c}{InvFels tree $\tau=\tau^*$, $t^*=\{x^*,y^*,x^*,y^*,y^*\}$}\\
    \hline
$\siteSplit_j$  & $p_{\siteSplit_j}$ &$8\cdot\Pr(\siteSplitRV=\siteSplit_j \mid \tau,t)$\\
    \hline
    $\emptyset$ & $p_{\emptyset}$   &$1+(x^*)^2+(y^*)^2+4x^*(y^*)^2+(x^*)^2(y^*)^2$\\
    $\{1\}$     & $p_{1}$           &$1-(x^*)^2+(y^*)^2-(x^*)^2(y^*)^2$\\
    $\{2\}$     & $p_{2}$           &$1+(x^*)^2-(y^*)^2-(x^*)^2(y^*)^2$\\
    $\{3\}$     & $p_{3}$           &$1-(x^*)^2+(y^*)^2-(x^*)^2(y^*)^2$\\
    $\{1,2,3\}$ & $p_{123}$         &$1+(x^*)^2-(y^*)^2-(x^*)^2(y^*)^2$\\
    $\{1,2\}$   & $p_{12}$          &$1-(x^*)^2-(y^*)^2+(x^*)^2(y^*)^2$\\
    $\{2,3\}$   & $p_{23}$          &$1-(x^*)^2-(y^*)^2+(x^*)^2(y^*)^2$\\
    $\{1,3\}$   & $p_{13}$          &$1+(x^*)^2+(y^*)^2-4x^*(y^*)^2+(x^*)^2(y^*)^2$\\
    \hline
    \multicolumn{3}{c}{InvFels tree $\tau=\tau_1$, $t=\{x_1,y_1,x_2,y_2,w\}$}\\
    \hline
$\siteSplit_j$  & $p_{\siteSplit_j}$ &$8\cdot\Pr(\siteSplitRV=\siteSplit_j \mid \tau,t)$\\
    \hline
    $\emptyset$ & $p_{\emptyset}$   &$1 + x_1x_2 +  y_1y_2 +  w[  x_1 + x_2][  y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{1\}$     & $p_{1}$           &$1 - x_1x_2 +  y_1y_2 +  w[- x_1 + x_2][  y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{2\}$     & $p_{2}$           &$1 + x_1x_2 -  y_1y_2 +  w[  x_1 + x_2][- y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{3\}$     & $p_{3}$           &$1 - x_1x_2 +  y_1y_2 +  w[  x_1 - x_2][  y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{1,2,3\}$ & $p_{123}$         &$1 + x_1x_2 -  y_1y_2 +  w[  x_1 + x_2][  y_1 - y_2] - x_1y_1x_2y_2$\\
    $\{1,2\}$   & $p_{12}$          &$1 - x_1x_2 -  y_1y_2 +  w[- x_1 + x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{2,3\}$   & $p_{23}$          &$1 - x_1x_2 -  y_1y_2 +  w[  x_1 - x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{1,3\}$   & $p_{13}$          &$1 + x_1x_2 +  y_1y_2 +  w[- x_1 - x_2][  y_1 + y_2] + x_1y_1x_2y_2$\\
    \hline
    \multicolumn{3}{c}{Felsenstein tree $\tau=\tau_2$, $t=\{x_1,y_1,x_2,y_2,w\}$}\\
    \hline
$\siteSplit_j$  & $p_{\siteSplit_j}$ &$8\cdot\Pr(\siteSplitRV=\siteSplit_j \mid \tau,t)$\\
    \hline
    $\emptyset$ & $p_{\emptyset}$   &$1 + x_1y_1 +  x_2y_2 +  w[  x_1 + y_1][  x_2 + y_2] + x_1y_1x_2y_2$\\
    $\{1\}$     & $p_{1}$           &$1 - x_1y_1 +  x_2y_2 +  w[ -x_1 + y_1][  x_2 + y_2] - x_1y_1x_2y_2$\\
    $\{2\}$     & $p_{2}$           &$1 - x_1y_1 +  x_2y_2 +  w[  x_1 - y_1][  x_2 + y_2] - x_1y_1x_2y_2$\\
    $\{3\}$     & $p_{3}$           &$1 + x_1y_1 -  x_2y_2 +  w[  x_1 + y_1][ -x_2 + y_2] - x_1y_1x_2y_2$\\
    $\{1,2,3\}$ & $p_{123}$         &$1 + x_1y_1 -  x_2y_2 +  w[ -x_1 - y_1][ -x_2 + y_2] - x_1y_1x_2y_2$\\
    $\{1,2\}$   & $p_{12}$          &$1 + x_1y_1 +  x_2y_2 +  w[ -x_1 - y_1][  x_2 + y_2] + x_1y_1x_2y_2$\\
    $\{2,3\}$   & $p_{23}$          &$1 - x_1y_1 -  x_2y_2 +  w[  x_1 - y_1][ -x_2 + y_2] + x_1y_1x_2y_2$\\
    $\{1,3\}$   & $p_{13}$          &$1 - x_1y_1 -  x_2y_2 +  w[ -x_1 + y_1][ -x_2 + y_2] + x_1y_1x_2y_2$\\
    \hline
\end{tabular}
\caption{Site split probabilities $p_{\siteSplit_j}$ on the true InvFels tree $\tau^*$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, and on the InvFels tree $\tau_1$ and Felsenstein tree $\tau_2$ with $t=\{x_1,y_1,x_2,y_2,w\}$ obtained using the Hadamard transform.
%EM "Multiplied by 1/8"... I think this should be "need to be multiplied by 1/8" or "All values have been divided by 1/8."
%das: added it to the table headings, too
All values need to be multiplied by $1/8$ to be probabilities.}
\label{tab:gen-sitepatprob}
\end{table}

@@

$\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$               \\
&$\{2\}^*$&$(1-x_1)(1-y_1)(1-x_2)(1-y_2)(1-w)$             \\
&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$             \\
\hline
\end{tabular}
\caption{
Likelihood calculations for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
All values need to be multiplied by $1/32$ to be probabilities.
Ancestral states with $^*$ are never maximal provided parameters are in $(0,1]$.
By combinations of $\ancestralSplit_k$, there are $3^5\cdot 4^2=3,888$ possible forms for the likelihood.
}
\label{tab:farris_likelihoods}
\end{table}

@@

$\{1,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
&&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
&&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\
\hline
\end{tabular}
\caption{
Likelihood calculations on the InvFels tree $\tau_1$.
Due to the symmetry of the likelihood, WLOG we assume $x_2 \ge x_1$ and $y_2 \ge y_1$ and maximize over ancestral state splits to reduce the number of possible functional forms to consider.
All values need to be multiplied by $1/32$ to be probabilities.
Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
Because in 4 cases there are 3 possibilities for $\xi_j$, there are $3^4=81$ possible forms for the likelihood.
}
\label{tab:likelihoods}
\end{table}

@@

\begin{proof}
We proceed by analyzing the gradients of each component of the likelihood function.
We first focus on the marginal log-likelihood
\[
\shannonDivergence_{\tau^*,t^*}(\tau_1,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t).
\]
By Gibbs' inequality, this function attains a unique maximum when
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)
\]
for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
We show that this implies $t=t^*$.
Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the marginal log-likelihood, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
Referring to the top of Table~\ref{tab:gen-sitepatprob}, we set up simultaneous equations as follows.
The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
\[
2-2(x^*)^2 = 2-2x^2
\]
and for $\{2\}$ and $\{1,2\}$ yields
\[
2-2(y^*)^2 = 2-2y^2.
\]
implying $x=x^*$ and $y=y^*$.
Setting these parameters to their optima, the difference of generating probabilities for site splits $\emptyset$ and $\{1,3\}$ yields
\[
8x^*(y^*)^2 = 8x^*y^*w
\]
implying $w=y^*$.
Thus the unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$, and consequently
\[
\nabla_t \shannonDivergence_{\tau^*,t^*}(\tau_1,t) \mid_{t=t^*} = 0.
\]
We now show the gradient of the marginal log-likelihood is a decreasing function on $(0,1]$ in each variable.
%EM I think you mean the divergence term as a function of x_1 is proportional to...
%das: removed "proportional" terminology abuse
The marginal log-likelihood is
\begin{align*}
%EM This is a little funky-- I think this is a definition of D(x_1)?
%das: made more explicit
\shannonDivergence(x_1) = \sum_j p_{\siteSplit_j} \log(c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1) - \log 8
\end{align*}
where $c_{x_1,\siteSplit_j}$ and $a_{x_1,\siteSplit_j}$ are functions of $y_1,x_2,y_2,$ and $w$.
For example, reading from Table~\ref{tab:gen-sitepatprob},
\[
c_{x_1,\emptyset} = 1+y_1y_2+wx_2y_1+wx_2y_2
\]
and
\[
a_{x_1,\emptyset} = x_2+wy_1+wy_2+x_2y_1y_2
\]
with similar definitions for the remaining site splits.
Consider the derivative with respect to $x_1$
\begin{align*}
\frac{d}{dx_1} \shannonDivergence(x_1) = \sum_j \frac{p_{\siteSplit_j}a_{x_1,\siteSplit_j}}{c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1}
\end{align*}
and focus on the $j$th term in the summation
\[
f_j(x_1) := \frac{p_{\siteSplit_j}a_{x_1,\siteSplit_j}}{c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1}.
\]
Recall that
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\}) = \frac{1}{8} (c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1),
\]
which, being a probability, is greater than or equal to zero.
Thus, regardless of $a_{x_1,\siteSplit_j} \lessgtr 0$, $f_j$ is a decreasing function of $x_1$ as $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 \ge 0$.
The remaining variables follow similar arguments.

\begin{table}
\centering
\begin{tabular}{|lll|l|}
\hline
$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau_1, t)$ & $\xi_j$ & $32\cdot\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau_1,t)$\\
\hline
$\emptyset$&$\{\emptyset\}$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{1\}$    &$\{\emptyset\}$&$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{2\}$    &$\{\emptyset\}$&$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{3\}$    &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,2,3\}$&$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,2\}$  &$\{\emptyset\}$&$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{2,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
\hline
\end{tabular}
\caption{
The maximal likelihood values on the InvFels tree $\tau_1$ assuming $\emptyset$ is the most likely ancestral state split for each site split.
All values need to be multiplied by $1/32$ to be probabilities.}
\label{tab:likelihoods-restricted}
\end{table}

We now focus on the conditional log-likelihood.
Assume $\emptyset$ is the most likely ancestral state split for all site splits.
In this case, the conditional log-likelihood has no ambiguity, and has terms given in Table~\ref{tab:likelihoods-restricted}.
We focus on the derivative with respect to $x_1$, aggregating all terms involving $x_2,y_1,y_2,$ and $w$ into a constant term.
The conditional log-likelihood
\[
\ell^C_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi_0) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV = \siteSplit_j, \tau_1, t),
\]
using Tables~\ref{tab:gen-sitepatprob} and~\ref{tab:likelihoods-restricted}, is
\begin{align*}
%EM Can you help me understand this? Where did the other terms from \log \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV = \siteSplit_j, \tau_1, t) go? They don't disappear with the proportionality-- they will be additive terms, not multiplicative. Or perhaps there's a bunch of cancellation? If the latter, let me know and I'll pull out my paper and pad.
%das: hope this is better?
\ell^C(x_1; \boldsymbol\xi_0) &= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
&\qquad + \left(\frac{1}{2}+\frac{1}{2}(y^*)^2\right)\log(1+y_1) + \left(\frac{1}{2}-\frac{1}{2}(y^*)^2\right)\log(1-y_1)\\
&\qquad + \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_2) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_2)\\
&\qquad + \log(1+y_2) + \log(1+w) - \log 32\\
&:= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
&\qquad + C_{x_1}(x_2,y_1,y_2,w).
\end{align*}
where $C_{x_1}$ is a function of $y_1,x_2,y_2,$ and $w$.
The derivative with respect to $x_1$ is
\[
\frac{d}{dx_1} \ell^C(x_1; \boldsymbol\xi_0) = \frac{-x_1+x^*(y^*)^2}{1-x_1^2}
\]
which is negative for all $x_1 \ge x^*$.
This same process applied to the remaining variables shows the gradient of the full likelihood is negative for $x_1,x_2 \ge x^*$ and $y_1 \ge y^*$ and positive for $y_2,w \ge y^*$, yielding the constraints
\[
\hat{x}_1,\hat{x}_2 \in (0, x^*], \ \hat{y}_1 \in (0, y^*], \text{and}\ \hat{y}_2,\hat{w} \in [y^*, 1].
\]
\end{proof}

@@

%EM "Proportional" thing again.
%das: got rid of all the proportional talk and just explicitly wrote out the likelihood

@@

We analyze the derivative with respect to $w$, so we absorb all terms of only $x_1,y_1,x_2,$ and $y_2$ into a constant that we ignore.

@@

Again, reading from Table~\ref{tab:gen-sitepatprob},

@@

%EM The conditional term is \log(1+w), right? If so, it'd be easier to have that second rather than first so that it corresponds to the order in eq:log_likelihood_simplified
%das: done

@@

        &:= \left(\sum_j p_{\siteSplit_j} \log(1 + b_{w,\siteSplit_j}w)\right) + \log(1+w) + C^*_w(x_1,y_1,x_2,y_2)

@@

% Hm, we didn't do a trick quite like this with the division bit.
%das: wrote out more explicitly

@@

This function has derivative with respect to $w$ as

@@

%%EM the negative sign                                                             V here doesn't seem right. We get the negative sign from going from p13 to the empty set.
%\ell'(w) &= \frac{1}{1+w} + \frac{p_{\emptyset}b_{\emptyset}}{1 + b_{\emptyset}w} - \frac{p_{13}b_{\emptyset}}{1 - b_{\emptyset}w} + \frac{p_{2}b_{2}}{1 + b_{2}w} - \frac{p_{2}b_{2}}{1 - b_{2}w} \\

>>>>

\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|}
    \multicolumn{3}{c}{InvFels tree $\tau=\tau^*$, $t^*=\{x^*,y^*,x^*,y^*,y^*\}$}\\
    \hline
$\siteSplit_j$  & $p_{\siteSplit_j}$ &$8\cdot\Pr(\siteSplitRV=\siteSplit_j \mid \tau,t)$\\
    \hline
    $\emptyset$ & $p_{\emptyset}$   &$1+(x^*)^2+(y^*)^2+4x^*(y^*)^2+(x^*)^2(y^*)^2$\\
    $\{1\}$     & $p_{1}$           &$1-(x^*)^2+(y^*)^2-(x^*)^2(y^*)^2$\\
    $\{2\}$     & $p_{2}$           &$1+(x^*)^2-(y^*)^2-(x^*)^2(y^*)^2$\\
    $\{3\}$     & $p_{3}$           &$1-(x^*)^2+(y^*)^2-(x^*)^2(y^*)^2$\\
    $\{1,2,3\}$ & $p_{123}$         &$1+(x^*)^2-(y^*)^2-(x^*)^2(y^*)^2$\\
    $\{1,2\}$   & $p_{12}$          &$1-(x^*)^2-(y^*)^2+(x^*)^2(y^*)^2$\\
    $\{2,3\}$   & $p_{23}$          &$1-(x^*)^2-(y^*)^2+(x^*)^2(y^*)^2$\\
    $\{1,3\}$   & $p_{13}$          &$1+(x^*)^2+(y^*)^2-4x^*(y^*)^2+(x^*)^2(y^*)^2$\\
    \hline
    \multicolumn{3}{c}{InvFels tree $\tau=\tau_1$, $t=\{x_1,y_1,x_2,y_2,w\}$}\\
    \hline
$\siteSplit_j$  & $p_{\siteSplit_j}$ &$8\cdot\Pr(\siteSplitRV=\siteSplit_j \mid \tau,t)$\\
    \hline
    $\emptyset$ & $p_{\emptyset}$   &$1 + x_1x_2 +  y_1y_2 +  w[  x_1 + x_2][  y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{1\}$     & $p_{1}$           &$1 - x_1x_2 +  y_1y_2 +  w[- x_1 + x_2][  y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{2\}$     & $p_{2}$           &$1 + x_1x_2 -  y_1y_2 +  w[  x_1 + x_2][- y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{3\}$     & $p_{3}$           &$1 - x_1x_2 +  y_1y_2 +  w[  x_1 - x_2][  y_1 + y_2] - x_1y_1x_2y_2$\\
    $\{1,2,3\}$ & $p_{123}$         &$1 + x_1x_2 -  y_1y_2 +  w[  x_1 + x_2][  y_1 - y_2] - x_1y_1x_2y_2$\\
    $\{1,2\}$   & $p_{12}$          &$1 - x_1x_2 -  y_1y_2 +  w[- x_1 + x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{2,3\}$   & $p_{23}$          &$1 - x_1x_2 -  y_1y_2 +  w[  x_1 - x_2][- y_1 + y_2] + x_1y_1x_2y_2$\\
    $\{1,3\}$   & $p_{13}$          &$1 + x_1x_2 +  y_1y_2 +  w[- x_1 - x_2][  y_1 + y_2] + x_1y_1x_2y_2$\\
    \hline
    \multicolumn{3}{c}{Felsenstein tree $\tau=\tau_2$, $t=\{x_1,y_1,x_2,y_2,w\}$}\\
    \hline
$\siteSplit_j$  & $p_{\siteSplit_j}$ &$8\cdot\Pr(\siteSplitRV=\siteSplit_j \mid \tau,t)$\\
    \hline
    $\emptyset$ & $p_{\emptyset}$   &$1 + x_1y_1 +  x_2y_2 +  w[  x_1 + y_1][  x_2 + y_2] + x_1y_1x_2y_2$\\
    $\{1\}$     & $p_{1}$           &$1 - x_1y_1 +  x_2y_2 +  w[ -x_1 + y_1][  x_2 + y_2] - x_1y_1x_2y_2$\\
    $\{2\}$     & $p_{2}$           &$1 - x_1y_1 +  x_2y_2 +  w[  x_1 - y_1][  x_2 + y_2] - x_1y_1x_2y_2$\\
    $\{3\}$     & $p_{3}$           &$1 + x_1y_1 -  x_2y_2 +  w[  x_1 + y_1][ -x_2 + y_2] - x_1y_1x_2y_2$\\
    $\{1,2,3\}$ & $p_{123}$         &$1 + x_1y_1 -  x_2y_2 +  w[ -x_1 - y_1][ -x_2 + y_2] - x_1y_1x_2y_2$\\
    $\{1,2\}$   & $p_{12}$          &$1 + x_1y_1 +  x_2y_2 +  w[ -x_1 - y_1][  x_2 + y_2] + x_1y_1x_2y_2$\\
    $\{2,3\}$   & $p_{23}$          &$1 - x_1y_1 -  x_2y_2 +  w[  x_1 - y_1][ -x_2 + y_2] + x_1y_1x_2y_2$\\
    $\{1,3\}$   & $p_{13}$          &$1 - x_1y_1 -  x_2y_2 +  w[ -x_1 + y_1][ -x_2 + y_2] + x_1y_1x_2y_2$\\
    \hline
\end{tabular}
\caption{8 times the site split probabilities $p_{\siteSplit_j}$ on the true InvFels tree $\tau^*$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$, and on the InvFels tree $\tau_1$ and Felsenstein tree $\tau_2$ with $t=\{x_1,y_1,x_2,y_2,w\}$ obtained using the Hadamard transform.
}
\label{tab:gen-sitepatprob}
\end{table}

@@

$\{1,3\}$  &$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$               \\
&$\{2\}^*$&$(1-x_1)(1-y_1)(1-x_2)(1-y_2)(1-w)$             \\
&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$             \\
\hline
\end{tabular}
\caption{
%EM Here and below, it seems a bit funny to call these likelihoods, as the Xi aren't observed.
32 times likelihoods for all site splits $\siteSplit_j$ and ancestral state splits $\ancestralSplit_k$ of the InvFels tree $\tau_1$.
Ancestral states with $^*$ are never maximal provided parameters are in $(0,1]$.
By combinations of $\ancestralSplit_k$, there are $3^5\cdot 4^2=3,888$ possible forms for the likelihood.
}
\label{tab:farris_likelihoods}
\end{table}

@@

$\{1,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
&&$\{1\}$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1-w)$\\
&&$\{1,2\}$&$(1+x_1)(1-y_1)(1+x_2)(1-y_2)(1+w)$\\
\hline
\end{tabular}
\caption{
32 times likelihoods on the InvFels tree $\tau_1$.
Due to the symmetry of the likelihood, WLOG we assume $x_2 \ge x_1$ and $y_2 \ge y_1$ and maximize over ancestral state splits to reduce the number of possible functional forms to consider.
Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
Because in 4 cases there are 3 possibilities for $\xi_j$, there are $3^4=81$ possible forms for the likelihood.
}
\label{tab:likelihoods}
\end{table}

@@

\begin{proof}
We proceed by analyzing the gradients of each component of the likelihood function.
We first focus on the marginal log-likelihood
\[
\shannonDivergence_{\tau^*,t^*}(\tau_1,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t).
\]
By Gibbs' inequality, this function attains a unique maximum when
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)
\]
for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
We show that this implies $t=t^*$.
Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the marginal log-likelihood, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
Referring to the top of Table~\ref{tab:gen-sitepatprob}, we set up simultaneous equations as follows.
The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
\[
2-2(x^*)^2 = 2-2x^2
\]
and for $\{2\}$ and $\{1,2\}$ yields
\[
2-2(y^*)^2 = 2-2y^2.
\]
implying $x=x^*$ and $y=y^*$.
Setting these parameters to their optima, the difference of generating probabilities for site splits $\emptyset$ and $\{1,3\}$ yields
\[
8x^*(y^*)^2 = 8x^*y^*w
\]
implying $w=y^*$.
Thus the unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$, and consequently
\[
\nabla_t \shannonDivergence_{\tau^*,t^*}(\tau_1,t) \mid_{t=t^*} = 0.
\]
We now show that the gradient of the marginal log-likelihood is a decreasing function on $(0,1]$ in each variable.
The marginal log-likelihood as a function of $x_1$ is
\begin{align*}
\shannonDivergence(x_1) = \sum_j p_{\siteSplit_j} \log(c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1) - \log 8
\end{align*}
where $c_{x_1,\siteSplit_j}$ and $a_{x_1,\siteSplit_j}$ are functions of $y_1,x_2,y_2,$ and $w$.
For example, reading from Table~\ref{tab:gen-sitepatprob},
\[
c_{x_1,\emptyset} = 1+y_1y_2+wx_2y_1+wx_2y_2
\]
and
\[
a_{x_1,\emptyset} = x_2+wy_1+wy_2+x_2y_1y_2
\]
with similar definitions for the remaining site splits.
Consider the derivative with respect to $x_1$
\begin{align*}
\frac{d}{dx_1} \shannonDivergence(x_1) = \sum_j \frac{p_{\siteSplit_j}a_{x_1,\siteSplit_j}}{c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1}
\end{align*}
and focus on the $j$th term in the summation
\[
f_j(x_1) := \frac{p_{\siteSplit_j}a_{x_1,\siteSplit_j}}{c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1}.
\]
Recall that
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\}) = \frac{1}{8} (c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1),
\]
which, being a probability, is greater than or equal to zero.
Thus, regardless of $a_{x_1,\siteSplit_j} \lessgtr 0$, $f_j$ is a decreasing function of $x_1$ as $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 \ge 0$.
The remaining variables follow similar arguments.

\begin{table}
\centering
\begin{tabular}{|lll|l|}
\hline
$\siteSplit_j$ & $\ancestralSplitPartition_j(\tau_1, t)$ & $\xi_j$ & $32\cdot\Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau_1,t)$\\
\hline
$\emptyset$&$\{\emptyset\}$&$\emptyset$&$(1+x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{1\}$    &$\{\emptyset\}$&$\emptyset$&$(1-x_1)(1+y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{2\}$    &$\{\emptyset\}$&$\emptyset$&$(1+x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{3\}$    &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1+x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,2,3\}$&$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,2\}$  &$\{\emptyset\}$&$\emptyset$&$(1-x_1)(1-y_1)(1+x_2)(1+y_2)(1+w)$\\
$\{2,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1+x_1)(1-y_1)(1-x_2)(1+y_2)(1+w)$\\
$\{1,3\}$  &$\{\emptyset,\{1\},\{1,2\}\}$&$\emptyset$&$(1-x_1)(1+y_1)(1-x_2)(1+y_2)(1+w)$\\
\hline
\end{tabular}
\caption{
32 times the maximal likelihood values on the InvFels tree $\tau_1$ assuming $\emptyset$ is the most likely ancestral state split for each site split.
}
\label{tab:likelihoods-restricted}
\end{table}

We now focus on the conditional log-likelihood.
Assume $\emptyset$ is the most likely ancestral state split for all site splits.
In this case, the conditional log-likelihood has no ambiguity, and has terms given in Table~\ref{tab:likelihoods-restricted}.
We focus on the derivative with respect to $x_1$, aggregating all terms involving $x_2,y_1,y_2,$ and $w$ into a constant term.
The conditional log-likelihood
\[
\ell^C_{\tau^*,t^*}(\tau_1, t; \boldsymbol\xi_0) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\emptyset \mid \siteSplitRV = \siteSplit_j, \tau_1, t),
\]
using Tables~\ref{tab:gen-sitepatprob} and~\ref{tab:likelihoods-restricted}, is
\begin{align*}
\ell^C(x_1; \boldsymbol\xi_0) &= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
&\qquad + \left(\frac{1}{2}+\frac{1}{2}(y^*)^2\right)\log(1+y_1) + \left(\frac{1}{2}-\frac{1}{2}(y^*)^2\right)\log(1-y_1)\\
&\qquad + \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_2) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_2)\\
&\qquad + \log(1+y_2) + \log(1+w) - \log 32\\
&:= \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1)\\
&\qquad + C_{x_1}(x_2,y_1,y_2,w).
\end{align*}
where $C_{x_1}$ is a function of $y_1,x_2,y_2,$ and $w$.
The derivative with respect to $x_1$ is
\[
\frac{d}{dx_1} \ell^C(x_1; \boldsymbol\xi_0) = \frac{-x_1+x^*(y^*)^2}{1-x_1^2}
\]
which is negative for all $x_1 \ge x^*$.
This same process applied to the remaining variables shows the gradient of the full likelihood is negative for $x_1,x_2 \ge x^*$ and $y_1 \ge y^*$ and positive for $y_2,w \ge y^*$, yielding the constraints
\[
\hat{x}_1,\hat{x}_2 \in (0, x^*], \ \hat{y}_1 \in (0, y^*], \text{and}\ \hat{y}_2,\hat{w} \in [y^*, 1].
\]
\end{proof}

@@

@@

We analyze the derivative with respect to $w$, so we absorb all terms of only $x_1,y_1,x_2,$ and $y_2$ into a constant that disappears upon taking the derivative.

@@

Using the middle panel of Table~\ref{tab:gen-sitepatprob}, we obtain

@@

@@

        &=: \left(\sum_j p_{\siteSplit_j} \log(1 + b_{w,\siteSplit_j}w)\right) + \log(1+w) + C^*_w(x_1,y_1,x_2,y_2)

@@

@@

The derivative with respect to $w$ of this function is

@@
